Interpretability and Fairness in Machine Learning: A Formal Methods Approach

### 2023-05-03 doctoral_consortium/main.tex
cr attempt 1

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness~\cite{ghosh2020classification}. Motivated by satisfying $ M $ out of $ N $ literals, we consider relaxed definitions of OR/AND operators in logic, which allows exceptions in the construction of a clause and in the selection of clauses in a rule. Based on these relaxed definitions, we introduce relaxed logical classification rules, which are inspired by the use of checklists in the medical domain and Boolean cardinality constraints. These rules generalize widely used rule representations, such as CNF, DNF, and decision sets. However, the combinatorial structure of these rules results in exponential succinctness, and na\"ive learning techniques are computationally expensive. To overcome this issue, we propose an incremental mini-batch learning procedure, called $\mathsf{CRR}$, which leverages advances in MILP solvers to efficiently learn such rules. Our experimental analysis shows that $\mathsf{CRR}$ can generate more accurate and sparser classification rules compared to alternative rule-based classifiers.

@@

\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness \textit{essentially beyond sensitive features}. Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias. To understand the sources of bias, it's necessary to determine \textit{which factors contribute how much to the bias of a classifier on a dataset}. We use a feature-attribution approach to explain the sources of bias, which relates the influences of input features to the resulting bias of the classifier. We formalize the Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}. We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis. In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions. In future, we aim to develop improved fairness enhancing and fairness attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.

>>>>

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness~\cite{ghosh2020classification}. Motivated by satisfying $ M $ out of $ N $ literals, we consider relaxed definitions of OR/AND operators in logic, which allows exceptions in the construction of a clause and in the selection of clauses in a rule. Based on these relaxed definitions, we introduce relaxed logical classification rules, which are inspired by the use of checklists in the medical domain and Boolean cardinality constraints. These rules generalize widely used rule representations, such as CNF, DNF, and decision sets. While the combinatorial structure of these rules results in exponential succinctness, na\"ive learning techniques are computationally expensive. To overcome this issue, we propose an incremental mini-batch learning procedure, called $\mathsf{CRR}$, which leverages advances in MILP solvers to efficiently learn such rules. Our experimental analysis shows that $\mathsf{CRR}$ can generate more accurate and sparser classification rules compared to alternative rule-based classifiers.

@@

\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness, \textit{essentially beyond sensitive features}. Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias. To understand the sources of bias, it's necessary to determine \textit{which factors contribute how much to the bias of a classifier on a dataset}. We use a feature-attribution approach to explain the sources of bias, which relates the influences of input features to the resulting bias of the classifier. We formalize Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}. We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis. In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions. In future, we aim to develop improved fairness enhancing and fairness attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.
